Theoretical aspects in gravity modelling

### 2018-04-26 manuscript/letter.tex
Including freeair and disturbance comparison

Other important reference system is a geocentric system of 
geodetic coordinates, which is defined by the reference ellipsoid 
used in the Normal Earth model \citep{heiskanen-moritz1967, soler1976, 
torge2012, bouman_etal2013}. 
In this coordinate system, the position of a point 
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Fig. \ref{fig:GCS-GGS}).
For convenience, we call this system Geocentric Geodetic System
(GGS). 
At a given point $(h, \varphi, \lambda)$, there are three 
mutually-orthogonal unit vectors (Fig. \ref{fig:GCS-GGS}) given by \citep{soler1976}:
\begin{equation}
\hat{\mitbf{u}} = 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix}
\hat{\mitbf{v}} = 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix}
\hat{\mitbf{w}} = 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix} .
\label{eq:unit-vectors}
\end{equation}
The required equations to convert coordinates $(h, \varphi, 
\lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to 
the GCS (Fig. \ref{fig:GCS-GGS}) and vice versa can be easily found in 
the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012, 
bouman_etal2013}.

@@

Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the 
gravity vector (corrected from non-gravitational effects due to vehicle 
motion and time variations such as Earth tides, instrumental drift and 
barometric pressure changes, for example) and the normal gravity vector 
at a point $P$. 
In this case, the gravity vector $\mitbf{g}_{P}$ represents the 
gradient of a scalar potential called \textit{gravity potential}, 
which is the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ represents the 
gradient of a scalar potential called \textit{normal potential}, 
which is also the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
By definition, the centrifugal part of the normal potential is equal 
to that of the gravity potential.
Then, the both gravity vector and normal gravity vectors are represented 
by the sum of two vectors: the gradient of a gravitational potential and
the gradient of a centrifugal potential. Notice that, as the centrifugal 
part of the gravity potential and normal potential are the same, the
centrifugal part of the gravity vector and normal gravity vector are
also equal to each other.

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
As the centrifugal part of the gravity vector and the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the (unknown) internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

@@

The gravity anomaly is the commonly used quantity in applied 
geophysics. It is defined as the difference between the 
gravity at a point $Q^{\prime}$ on the Geoid (a particular 
equipotential surface of the gravity potential) and the normal gravity 
at a point $Q$ on the ellipsoid surface,
both at the same geodetic latitude and longitude (Fig. \ref{fig:surfaces}).
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have computed the upward 
continuation of gravity anomalies. All these authors have implicitly 
considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that can be continued upward.

@@

@@

>>>>

Another important reference system is a geocentric system of 
geodetic coordinates, which is defined by the reference ellipsoid 
used in the Normal Earth model \citep{heiskanen-moritz1967, soler1976, 
torge2012, bouman_etal2013}. 
In this coordinate system, the position of a point 
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Fig. \ref{fig:GCS-GGS}).
For convenience, we call this system Geocentric Geodetic System
(GGS). 
At a given point $(h, \varphi, \lambda)$, there are three 
mutually-orthogonal unit vectors (Fig. \ref{fig:GCS-GGS}) given by \citep{soler1976}:
\begin{equation}
\hat{\mitbf{u}} = 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix}
\hat{\mitbf{v}} = 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix}
\hat{\mitbf{w}} = 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix} .
\label{eq:unit-vectors}
\end{equation}
The required equations to convert coordinates $(h, \varphi, 
\lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to 
the GCS (Fig. \ref{fig:GCS-GGS}) and vice versa can be easily found in 
the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012, 
bouman_etal2013}.

@@

Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the 
gravity vector (corrected from non-gravitational effects due to vehicle 
motion and time variations such as Earth tides, instrumental drift and 
barometric pressure changes, for instance) and the normal gravity vector 
at a point $P$. 
In this case, the gravity vector $\mitbf{g}_{P}$ represents the 
gradient of a scalar potential called \textit{gravity potential}, 
which is the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ represents the 
gradient of a scalar potential called \textit{normal potential}, 
which is also the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
By definition, the centrifugal part of the normal potential is equal 
to that of the gravity potential.
Then, the both gravity vector and normal gravity vectors are represented 
by the sum of two vectors: the gradient of a gravitational potential and
the gradient of a centrifugal potential. Notice that, as the centrifugal 
part of the gravity potential and normal potential are the same, the
centrifugal part of the gravity vector and normal gravity vector are
also equal to each other.

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
As the centrifugal part of the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the (unknown) internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

@@

The gravity anomaly is the commonly used quantity in applied 
geophysics. At point $P$, it is defined as the difference between the 
gravity vector at a point $Q^{\prime}$ on the Geoid (a particular 
equipotential surface of the gravity potential) and the normal gravity 
vector at a point $Q$ on the ellipsoid surface, both at the same
geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
\begin{equation}
\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime} } - \mitbf{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}
Similarly as the gravity disturbance expression (eq.
\ref{eq:gravity-disturbance}), the gravity anomaly is given by
\begin{equation}
\Delta g_{P} = g_{Q^{\prime} } - \gamma_{Q} .
\label{eq:gravity-anomaly}
\end{equation}
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have. All these authors have
implicitly considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that is allowed to be upward continued.

@@

\section{Comparing Free-air anomaly with gravity disturbance}

By definition, for a small height $h$, the normal gravity at point
$P(h, \varphi, \lambda)$ can be expanded into a Taylor series 
in terms of $h$ \citep{heiskanen-moritz1967,
hofmann-wellenhof-moritz2005} as in
\begin{equation}
\gamma_{P} \approx \gamma_{Q} + \frac{\partial \gamma}{\partial h} h ,
\label{eq:gamma-expansion}
\end{equation}
where higher orders of $h$ were neglected. $h$ represents the
distance between points $P$ and $Q$ shown in (Fig. \ref{fig:heights}).
Therefore, substituting \ref{eq:gamma-expansion}'s result into
eq. \ref{eq:gravity-disturbance}, the gravity disturbance becomes
\begin{equation}
\delta g_{P} \approx g_{P} - (\gamma_{Q} + \frac{\partial \gamma}{\partial h} h) .
\label{eq:disturb-approx}
\end{equation}
According to \citet{heiskanen-moritz1967, blakely1996}, the relation
between gravity on the geoid and on the reference ellipsoid is
a similar Taylor expansion given by
\begin{equation}
g_{Q^{\prime}} = g_{P} - \frac{\partial \gamma}{\partial h} H
\label{eq:geod2phys_surf}
\end{equation}
where $H$ is defined as the orthometric height (Fig. \ref{fig:heights})
measured along the curved plumb line, starting from the Geoid
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
Thus, the Free-air anomaly $(\Delta g^{a})$, which corrects for the
decrease in gravity due to the elevation of the observation point $P$, 
can be written as
\begin{equation}
\Delta g^{a}_{P} \approx g_{P} - (\gamma_{Q} + \frac{\partial \gamma}{\partial h} H)
\label{eq:free-air}
\end{equation}
By inspection, one easily concludes that the difference is the first
order derivative term
\begin{equation}
\delta g_{P} - \Delta g^{a}_{P} \approx - \frac{\partial \gamma}{\partial h} (h-H) = - \frac{\partial \gamma}{\partial h} N ,
\label{eq:disturb-FA}
\end{equation}
where $N$ is the geoidal undulation (Fig. \ref{fig:heights}) and 
$\frac{\partial \gamma}{\partial h} = -0.3086$.  Hence, using a
mean value of $\tilde{N} = 120$m, the absolute value of the
difference in eq. \ref{eq:disturb-FA} is computed
\begin{equation}
\parallel \delta g_{P} - \Delta g^{a}_{P} \parallel \leq \parallel - \frac{\partial \gamma}{\partial h} \tilde{N} \parallel = 37.032
\label{eq:diff-constant}
\end{equation}

@@

\begin{figure}
    \includegraphics{figures/different-heights.png}
    \caption{Schematic representation of the geometric height
        $h$, orthometric height $H$ and geoidal undulation $N$.
        $H$ is curved because it follows the plumbline.}
  \label{fig:heights}
\end{figure}
